<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{rerddap introduction}
%\VignetteEncoding{UTF-8}
-->

```{r echo=FALSE}
library("knitr")
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines) == 1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(if (abs(lines[1]) > 1) more else NULL,
            x[lines],
            if (length(x) > lines[abs(length(lines))]) more else NULL
           )
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
)
```

rerddap introduction
====================

`rerddap` is a general purpose R client for working with ERDDAP servers. ERDDAP is a server built on top of OPenDAP, which serves some NOAA data. You can get gridded data ([griddap](https://upwell.pfeg.noaa.gov/erddap/griddap/documentation.html)), which lets you query from gridded datasets, or table data ([tabledap](https://upwell.pfeg.noaa.gov/erddap/tabledap/documentation.html)) which lets you query from tabular datasets. In terms of how we interface with them, there are similarties, but some differences too. We try to make a similar interface to both data types in `rerddap`.

## NetCDF

`rerddap` supports NetCDF format, and is the default when using the `griddap()` function. NetCDF is a binary file format, and will have a much smaller footprint on your disk than csv. The binary file format means it's harder to inspect, but the `ncdf4` package makes it easy to pull data out and write data back into a NetCDF file. Note the the file extension for NetCDF files is `.nc`. Whether you choose NetCDF or csv for small files won't make much of a difference, but will with large files.

## Caching

Data files downloaded are cached in a single hidden directory `~/.rerddap` on your machine. It's hidden so that you don't accidentally delete the data, but you can still easily delete the data if you like.

When you use `griddap()` or `tabledap()` functions, we construct a MD5 hash from the base URL, and any query parameters - this way each query is separately cached. Once we have the hash, we look in `~/.rerddap` for a matching hash. If there's a match we use that file on disk - if no match, we make a http request for the data to the ERDDAP server you specify.

## ERDDAP servers

You can get a data.frame of ERDDAP servers using the function `servers()`. Most I think serve some kind of NOAA data, but there are a few that aren't NOAA data.  If you know of more ERDDAP servers, send a pull request, or let us know.

## Install

Stable version from CRAN

```{r eval=FALSE}
install.packages("rerddap")
```

Or, the development version from GitHub

```{r eval=FALSE}
devtools::install_github("ropensci/rerddap")
```

```{r}
library("rerddap")
```

## Search

First, you likely want to search for data, specify either `griddadp` or `tabledap`

```{r}
ed_search(query = 'size', which = "table")
```

```{r}
ed_search(query = 'size', which = "grid")
```

## Information

Then you can get information on a single dataset

```{r output.lines=1:10}
info('whoi_62d0_9d64_c8ff')
```

## griddap (gridded) data

First, get information on a dataset to see time range, lat/long range, and variables.

```{r}
(out <- info('noaa_esrl_027d_0fb5_5d38'))
```

Then query for gridded data using the `griddap()` function

```{r}
(res <- griddap(out,
  time = c('2012-01-01', '2012-01-30'),
  latitude = c(21, 10),
  longitude = c(-80, -70)
))
```

The output of `griddap()` is a list that you can explore further. Get the summary

```{r output.lines=1:15}
res$summary
```

Get the dimension variables

```{r}
names(res$summary$dim)
```

Get the data.frame (beware: you may want to just look at the `head` of the data.frame if large)

```{r}
res$data
```

## tabledap (tabular) data

```{r output.lines=1:10}
(out <- info('erdCalCOFIfshsiz'))
```

```{r}
(dat <- tabledap(out, 'time>=2001-07-07', 'time<=2001-07-10', fields = c('longitude', 'latitude', 'fish_size', 'itis_tsn', 'scientific_name')))
```

Since both `griddap()` and `tabledap()` give back data.frame's, it's easy to do downstream manipulation. For example, we can use `dplyr` to filter, summarize, group, and sort:

```{r}
library("dplyr")
dat$fish_size <- as.numeric(dat$fish_size)
tbl_df(dat) %>%
  filter(fish_size > 30) %>%
  group_by(scientific_name) %>%
  summarise(mean_size = mean(fish_size)) %>%
  arrange(desc(mean_size))
```
